Robust Spectroscopic Inference with Imperfect Models

### 2014-09-13 tex/paper.tex
Draft for Sean

\begin{abstract}
We present a modular and extensible framework for forward-modelling spectra with synthetic models. Due to the nature of spectra, that adjacent pixels are not independent samples of the intrinsic flux distribution, it is common to result in residual spectra that exhibit some degree of correlation. Neglecting this covariance in the residual spectrum can result in a biased estimates of the parameters which describe the model spectrum. In the high signal-to noise, large spectral range limit common to stellar parameter estimation, the covariant structure of the residuals can bias the resulting parameter determinations. We present a likelihood function designed to account for this covariant structure, utilizing the machinery of Gaussian process covariance kernels to self-consistently identify and model the covariance. Specific to stellar parameter estimation, a common problem is the presence of spectral lines with an incorrect strength, often due to incorrect atomic constants or radiative treatment. We also present the development of a local covariance kernel that can track and self-consistently down-weight pathologically wrong spectra lines. Local kernels that track line weights provide a mechanism to learn corrections to spectral models, by fitting in a hierarchical manner. We demonstrate the application of this method, implemented as an open source code, to fit an echelle optical spectrum WASP-14, an F star, and a mid-infra spectrum of Gl51, an M5 star. 
\end{abstract}
\keywords{stars: fundamental parameters --- methods: data analysis --- methods: statistical}

@@

\begin{deluxetable}{lr@{ $\pm$ }l|r@{ $\pm$ }l}[!b]
\tablecaption{\label{table:Kurucz}Inferred Parameters for WASP-14}
\tablehead{\colhead{Parameter} & \multicolumn{2}{c}{{\sc CfA/Kurucz}} & \multicolumn{2}{c}{{\sc Phoenix}}}
\startdata
$T_{\rm eff}$ (K) & 6517     & 13                      &  6273   & 15        \\
$\log g$      & 4.29     & 0 (fixed)               &  4.29   & 0 (fixed) \\
$\Z$          & -0.27    & 0.01                    & -0.498  & 0.006     \\
$v \sin i$ ($\kms$)\tablenotemark{a}   & 4.29     & 0.05                    &  4.89   & 0.05      \\
$v_z$ ($\kms$)        & -4.62    & 0.02                    & -4.86   & 0.02      \\
$\log \Omega$\tablenotemark{b} & -12.7181 & 0.0004 & -19.678 & 0.001
\enddata
\tablenotetext{a}{Small differences in the rotational broadening between the two libraries are
expected, since they assume different micro- and macro-turbulence behavior in the stellar
atmospheres.}
\tablenotetext{b}{The {\sc CfA/Kurucz} spectra are scaled such that the average flux is unity, rather than given in units of flux at the stellar surface, so this value is an unknown multiple of $R^2/d^2$.  The inference on $\Omega$ from the {\sc Phoenix} library is in steradian units.}
\tablecomments{The quoted values represent the ``best-fit", the peak of the marginal posterior 
distributions for each parameter.  The quoted uncertainties correspond to the 68.3\%\ 
($\sim$1\,$\sigma$) confidence intervals.  Expanded lists of the inferred values and uncertainties 
for all nuisance parameters are available electronically.}
\end{deluxetable}

@@

\section{Interpolation Errors}
The interpolation of a model spectrum from a synthetic library of spectra is a conceptually straightforward process; however, in practice there may be several subtle effects that require consideration, depending on the application. Spectra are generally stored in a synthetic library as 1-D arrays of flux values, sampled at specific wavelengths at high resolution (typically R$\gtrsim$100,000). Each spectrum has several fundamental parameters which describe the model that generated it. To use a concrete example, the stellar libraries used in this paper have three parameters, $T_\textrm{eff}$, $\log_{10} g$, and $\Z$, each of which constitutes an axis in the synthetic library. Therefore, the full synthetic grid can be stored as a 4-D array, where the first three axes are the stellar parameters and the 4th axis describes the wavelength. The value stored in the array is the flux of the spectrum described by current value of the four axes.

Spectral line formation is a non-linear process, with the shape of the line determined by complicated radiative synthesis, which is a function of the stellar parameters. Interpolating a \emph{physical} spectrum short-cuts this non-linear process and attempts to predict the behavior of the spectral line based upon the behavior of nearby spectra in the grid. While this approximation is generally accurate enough for our applications, for high signal-to-noise applications it becomes necessary to account for any inaccuracies introduced by the interpolation. In practice, there are two factors which influence the choice of interpolation method. First, for a small percentage of pixels, the flux value will not vary monotonically with change in fundamental parameters. Second, visual inspection of the pixel values as a function of fundamental parameters makes it apparent that the spectra are not Nyquist sampled in parameter space. This means that synthetic grid does is not dense enough to losslessly reconstruct any synthetic spectrum. However, this fact is not surprising given the computational complexity of stellar synthesis. Given this reality, we make the most conservative assumption about the behavior of the pixels between grid points and decide to use a linear interpolation scheme. Using a more advanced interpolation method, such as a spline or other band-limited interpolation scheme, would not remedy the fact that the grid itself is not Nyquist sampled. To account for the loss of spectral information during the interpolation process, we build a model of interpolation error and incorporate this into the likelihood function (Eqn.~\ref{eqn:lnlikelihood}).

We empirically determine the interpolation error by performing a series of ``drop-out'' interpolation tests, whereby we remove a spectrum from the grid, generate an interpolated spectrum from the nearest two grid points, and compare the difference between the two. For example, consider the {\sc Phoenix} grid used for the WASP-14 parameter estimation. In this parameter space, the grid is sampled every 100K in $T_\textrm{eff}$ and every 0.5 dex in $\log_{10} g$ and $\Z$. Therefore, for each spectrum in the synthetic grid, there are also three error spectra associated with it, one for each axis of stellar parameters. For example, consider a spectrum with stellar parameters $T_\textrm{eff} = $ 6000K, $\log_{10} g=$4.0 dex, and $\Z = $ -0.5 dex. A temperature ``error spectrum'' would be formed by subtracting this true spectrum from the spectrum formed by the average of the $T_\textrm{eff} = $ 6100K spectrum and the $T_\textrm{eff} = $ 5900K spectrum, with $\log_{10} g$ and $\Z$ held constant. Likewise, the $\log_{10} g$ and $\Z$ error spectra are created by performing the same operation over the $\log_{10} g$ and $\Z$ axes of the grid (see Figure~\ref{fig:interp_errors}). Because the error spectra are determined over an interpolation distance twice that (e.g., 100K) of the actual distance used (e.g., 50K), we reduce the amplitude of the error spectra by two. Drop-out tests at larger interpolation distances (e.g., 200K and 300K) show that the amplitude of the error spectrum grows approximately linearly with interpolation distance.

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_errors.pdf}
  \figcaption{(\emph{top}) A small range of the ``true'' {\sc Phoenix} spectrum at $T_\textrm{eff} = $ 6000K, $\log_{10} g=$4.0 dex, and $\Z = $ -0.5 dex, shown with the drop-out interpolation spectra along each of the three parameter axes. These interpolation error spectra are determined over an interpolation distance twice that (e.g., 100K) of the actual distance used in practice. Typically, the interpolated spectra closely match the ``true'' spectrum. (\emph{bottom}) The three ``error spectra'', or difference between the interpolated spectrum and the true spectrum, relative to the true spectrum value. The shaded region shows the total change in flux between the two spectra used for interpolation, relative to the true spectrum. Note that the error spectrum is generally much smaller than the typical change in flux. For the few points where this is not true (e.g., $\log_{10} g$ near 5161.5\AA) this implies a non-monotonic change in flux between grid points. Therefore, it is best to conservatively use the error spectrum, which protects against assumptions of monotonicity.}
  \label{fig:interp_errors}
\end{center}
\end{figure}

Generally, the interpolation error is small: typically no larger than 2\% of the flux and usually--but not always--much smaller than the total flux difference between spectra at adjacent grid points (see Figure~\ref{fig:interp_errors}). This test gives us confidence that it is possible to generate an interpolated spectra, and that inference on the stellar parameters may be drawn at a substantially finer level that the coarsity of the grid spacing itself, especially given the large number of possible lines to fit. However, while the interpolation error is small, because all spectral lines have roughly the same shape, it is often true that the error adds \emph{coherently}. When the signal to noise of the spectrum is particularly large, this coherent trend can contribute substantially by diminishing the likelihood of an interpolated spectrum, since the lines in the spectrum are less ``physical'' in shape than that of a spectrum directly on a gridpoint, which requires minor or no interpolation. Therefore in order to deliver the proper likelihood for interpolated spectra it is necessary to quantify the degree of interpolation error and add this into the covariance matrix.

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_grid.pdf}
  \figcaption{Tri-linear interpolation on a grid cell the shape of a rectangular prism.}
  \label{fig:interp_grid}
\end{center}
\end{figure}

Because the interpolation error is highly dependent on whether the wavelength is on the continuum or on a spectral line, we must model the covariance of the errors. For example, all pixels at the continuum will tend to act similarly, regardless of where on the spectrum the pixels fall. Any pixel that falls on the blue side of a spectral line will tend to exhibit a similar error behaviour regardless of which spectral line, as long as that line was changing in a similar manner. The elements of the interpolation covariance matrix $\mathsf{C}_{ij}^\textrm{int}$ are determined empirically from the drop-out spectra, and are a function of the degree of the interpolation. Consider the manner in which a typical spectrum is tri-linearly interpolated from the synthetic grid (Figure~\ref{fig:interp_grid}). The interpolated spectrum is a weighted average of the neighboring 8 spectra, with weights $d_l$ based upon the distance from the grid points. To generate a composite covariance matrix which reflects the interpolation error present for the interpolated spectra, we must also combine the error spectra. Because the error is covariant, we must first multiply and then add the error spectra, in order to preserve the covariance. A simple sum of error spectra would destroy any information about the covariance. For a specific axis $k$ (e.g., $T_\textrm{eff}$, $\log_{10} g$, or $\Z$), we represent the average covariance matrix as a composite of all of the error spectra of this species
\begin{equation}
  \mathsf{C}_{ijk}^\textrm{int} = \sum_{l=1}^8  d_{kl} \lambda_{ikl} \lambda_{jkl}
\end{equation}

@@

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_matrix.pdf}
  \figcaption{A typical interpolation error matrix for $T_\textrm{eff} = $ 6220K, $\log_{10} g=$4.2 dex, and $\Z = $ -0.4 dex.}
  \label{fig:interp_matrix}
\end{center}
\end{figure}

>>>>

\begin{abstract}
We present a modular and extensible framework for inference of model parameters using synthetic spectra. By the nature of continuously sampled spectra, when subtracting a synthetic model, adjacent pixels in the residual spectrum exhibit some degree of correlation. In the high signal-to noise, large spectral range limit common to stellar parameter estimation, the covariant structure of the residuals can bias the resulting parameter determinations. We present a likelihood function designed to account for this covariant structure, utilizing the machinery of Gaussian process covariance kernels to self-consistently identify and model the covariance. Additionally, a common problem specific to stellar parameter estimation is the presence of spectral lines with mismatched strength, often due to incorrect atomic constants or radiative treatment. To address this issue, we also present the development of a novel local covariance kernel that can identify and self-consistently down-weight pathologically wrong spectral line ``outliers.'' By fitting multiple spectra in a hierarchical manner, these local kernels provide a mechanism to learn empirical corrections to spectral models. We demonstrate the application of this method, implemented as a freely available open source code, by fitting an high resolution optical spectrum WASP-14, an F star, and a moderate resolution mid-infrared spectrum of Gl51, an M5 star. 
\end{abstract}
\keywords{stars: fundamental parameters --- methods: data analysis --- methods: statistical}

@@

\begin{deluxetable}{lr@{ $\pm$ }l|r@{ $\pm$ }l}[!b]
\tablecaption{\label{table:Kurucz}Inferred Parameters for WASP-14}
\tablehead{\colhead{Parameter} & \multicolumn{2}{c}{{\sc CfA/Kurucz}} & \multicolumn{2}{c}{{\sc Phoenix}}}
\startdata
$T_{\rm eff}$ (K) & 6523     & 14                      &  6234   & 17        \\
$\log g$      & 4.29     & 0 (fixed)               &  4.29   & 0 (fixed) \\
$\Z$          & -0.28    & 0.01                    & -0.52  & 0.02     \\
$v \sin i$ ($\kms$)\tablenotemark{a}   & 4.28     & 0.06                    &  4.89   & 0.06      \\
$v_z$ ($\kms$)        & -4.62    & 0.02                    & -4.86   & 0.02      \\
$\log \Omega$\tablenotemark{b} & -12.7180 & 0.0004 & -19.676 & 0.001
\enddata
\tablenotetext{a}{Small differences in the rotational broadening between the two libraries are
expected, since they assume different micro- and macro-turbulence behavior in the stellar
atmospheres.}
\tablenotetext{b}{The {\sc CfA/Kurucz} spectra are scaled such that the average flux is unity, rather than given in units of flux at the stellar surface, so this value is an unknown multiple of $R^2/d^2$.  The inference on $\Omega$ from the {\sc Phoenix} library is in steradian units.}
\tablecomments{The quoted values represent the ``best-fit", the peak of the marginal posterior 
distributions for each parameter.  The quoted uncertainties correspond to the 68.3\%\ 
($\sim$1\,$\sigma$) confidence intervals.  Expanded lists of the inferred values and uncertainties 
for all nuisance parameters are available electronically.}
\end{deluxetable}

@@

\section{Interpolation Errors}
The interpolation of a model spectrum from a synthetic library of spectra is a conceptually straightforward process; however, depending on the application, in practice there may be several subtle effects that require consideration. Spectra are generally stored in a synthetic library as 1-D arrays of flux values, sampled at specific wavelengths at high resolution (typically R$\gtrsim$100,000). Each spectrum has several fundamental parameters which describe the model that generated it. For example, the stellar libraries used in this paper have three parameters, $T_\textrm{eff}$, $\log_{10} g$, and $\Z$, each of which constitutes an axis in the synthetic library. Therefore, the full synthetic grid is stored as a 4-D array, where the first three axes are the stellar parameters and the 4th axis describes the wavelength. The value stored in the array is the flux of the spectrum described by the current value of the four axes.

Spectral line formation is a non-linear process: line shapes are determined by the complicated radiative transfer through the stellar atmosphere, the structure of which is iterated to convergence using a specific set of stellar parameters. Interpolating a spectrum short-cuts this non-linear process and attempts to predict the behavior of the spectral line based upon the behavior of nearby spectra in the grid. While this approximation is generally accurate enough for our applications, in the high signal-to-noise regime it becomes necessary to account for any inaccuracies introduced by the interpolation. In practice, there are two factors to consider when interpolating spectra. First, for a small percentage of pixels, the flux value will not vary monotonically with change in fundamental parameters. This informs the decision about which functional form of interpolation to use. Second, inspecting the flux values of a sample of pixels as a function of fundamental parameters makes it apparent that not all pixels of the spectrum are Nyquist sampled with regards to the stellar parameters. This unfortunately means that the synthetic grid is not sampled densely enough to losslessly reconstruct a synthetic spectrum. However, given the computational complexity of stellar synthesis, this fact is not surprising. In order to proceed, we make the most conservative assumption about the behavior of the pixels between grid points and decide to use a linear interpolation scheme. Using a more advanced interpolation method, such as a spline or other band-limited interpolation scheme, would not remedy the fact that the grid itself is not Nyquist sampled. To account for the loss of spectral information during the interpolation process, we build a model of interpolation error and incorporate this into the likelihood function (Eqn.~\ref{eqn:lnlikelihood}).

We empirically determine the interpolation error by performing a series of ``drop-out'' interpolation tests, whereby we remove a spectrum from the grid, generate an interpolated spectrum from the nearest two grid points, and compare the difference between the two. As a concrete example, the {\sc Phoenix} grid used for the WASP-14 parameter estimation is sampled every 100K in $T_\textrm{eff}$ and every 0.5 dex in $\log_{10} g$ and $\Z$. Therefore, for each spectrum in the synthetic grid, there are also three error spectra associated with it, one for each axis of stellar parameters. For example, consider a spectrum with stellar parameters $T_\textrm{eff} = $ 6000K, $\log_{10} g=$4.0 dex, and $\Z = $ -0.5 dex. A temperature ``error spectrum'' would be formed by subtracting this true spectrum from the spectrum formed by the average of the $T_\textrm{eff} = $ 6100K spectrum and the $T_\textrm{eff} = $ 5900K spectrum, with $\log_{10} g$ and $\Z$ held constant. Likewise, the $\log_{10} g$ and $\Z$ error spectra are created by performing the same operation over the $\log_{10} g$ and $\Z$ axes of the grid (see Figure~\ref{fig:interp_errors}). Because the error spectra are determined over an interpolation distance twice that (e.g., 100K) of the actual distance used (e.g., 50K), we reduce the amplitude of the error spectra by two. Drop-out tests at larger interpolation distances (e.g., 200K and 300K) show that the amplitude of the error spectrum grows approximately linearly with interpolation distance.

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_errors.pdf}
  \figcaption{(\emph{top}) A small range of the ``true'' {\sc Phoenix} spectrum at $T_\textrm{eff} = $ 6000K, $\log_{10} g=$4.0 dex, and $\Z = $ -0.5 dex, shown with the drop-out interpolation spectra along each of the three parameter axes. These interpolation error spectra are determined over an interpolation distance twice that (e.g., 100K) of the actual distance used while fitting. The interpolated spectra closely match the ``true'' spectrum. (\emph{bottom}) The three ``error spectra'', or difference between the interpolated spectrum and the true spectrum, relative to the true spectrum value. The shaded region shows the total change in flux between the two spectra used for interpolation, relative to the true spectrum. Note that the error spectrum is generally much smaller than the typical change in flux. For the few points where this is not true (e.g., $\log_{10} g$ near 5161.5\AA) this implies a non-monotonic change in flux between grid points. Therefore, it is most conservative to use the error spectrum in an interpolation error model, which accounts for this behavior and protects against assumptions of monotonicity.}
  \label{fig:interp_errors}
\end{center}
\end{figure}

Generally, the interpolation error is small: typically no larger than 2\% of the flux (smaller than the Poisson noise for moderate to high signal-to-noise spectra) and usually--but not always--much smaller than the total flux difference between spectra at adjacent grid points (see Figure~\ref{fig:interp_errors}). This test gives us confidence that it is possible to generate a sufficiently accurate interpolated spectrum, and that inference on the stellar parameters may be drawn at a substantially finer level that the coarsity of the grid spacing itself, especially given the large number of possible lines to fit. However, while the interpolation error is small, because all spectral lines have roughly the same shape, it is often true that the error adds \emph{coherently}. When the signal to noise of the spectrum is particularly high, this coherent trend can contribute substantially by diminishing the likelihood of an interpolated spectrum, since the lines in the spectrum are less ``physical'' in shape than that of a spectrum directly on a gridpoint, which requires minor or no interpolation. Therefore in order to deliver the proper likelihood for interpolated spectra it is necessary to quantify the degree of interpolation error and add this into the covariance matrix.

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_grid.pdf}
  \figcaption{Tri-linear interpolation on a grid cell the shape of a rectangular prism. I'm not sure if this figure is necessary, but I think it could help guide the reader. Perhaps by labelling the axis with temperature and so on. It is here as a prototype.}
  \label{fig:interp_grid}
\end{center}
\end{figure}

Because the interpolation error is highly dependent on whether the wavelength is on the continuum or on a spectral line, we must model the covariance of the errors. For example, all pixels at the continuum will tend to act similarly, regardless of where on the spectrum the pixels fall. All pixels that fall on the blue side of a spectral line will tend to exhibit a similar error behaviour regardless of which spectral line they belong to, as long as the general behavior of the spectral lines was similar. The elements of the interpolation covariance matrix $\mathsf{C}_{ij}^\textrm{int}$ are determined empirically from the drop-out spectra, and are a function of the degree of the interpolation. Consider the manner in which a typical spectrum is tri-linearly interpolated from the synthetic grid (Figure~\ref{fig:interp_grid}). The interpolated spectrum is a weighted average of the neighboring 8 spectra, with weights $d_l$ based upon the distance from the nearest grid points $P$. To generate a composite covariance matrix which reflects the interpolation error present for the interpolated spectra, we must also combine the error spectra. Because the error is covariant, we must first determine the empirical covariance matrix for each error spectrum and then add the covariance matrices. A simple sum of error spectra would destroy any information about the covariance. The empirical covariance matrix generated by a single error spectrum $l$, for a specific axis $k$ (e.g., $T_\textrm{eff}$, $\log_{10} g$, or $\Z$), is
\begin{equation}
  \mathsf{C}_{ijkl}^\textrm{int} = \lambda_{ikl} \lambda_{jkl}
\end{equation}
We represent the average covariance matrix as a composite of all of the error covariance matrices of this species
\begin{equation}
  \mathsf{C}_{ijk}^\textrm{int} = \sum_{l=1}^8  d_{l} \mathsf{C}_{ijl}^\textrm{int}
\end{equation}

@@

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/interp_matrix.pdf}
  \figcaption{A typical interpolation error matrix for $T_\textrm{eff} = $ 6220K, $\log_{10} g=$4.2 dex, and $\Z = $ -0.4 dex. Some elements of the covariance matrix may be negative.}
  \label{fig:interp_matrix}
\end{center}
\end{figure}
